metric-learn: Metric Learning Algorithms in Python

### 2019-08-07 metric-learn-jmlr.tex

Metric learning algorithms can be classified according to the form of data supervision that they need to learn the metric. \texttt{metric-learn} currently implements algorithms that fall into the following categories:
\begin{itemize}
\item \textbf{Supervised learners} learn from a standard labeled dataset and essentially aim to bring closer together points from the same class while spreading further away points from a different class. For instance, the class may be the identity of a person in an image (see Figure~\ref{fig:full}). 
% \todo[inline]{ces figures ne sont pas très vendeuses. je propose de faire une figure conceptuelle en prenant l'exemple des faces expliquant la différence entre le supervisé, les paires et les quadruplets. on peut s'inspirer de \url{http://cedric.cnam.fr/~thomen/papers/Law_ICCV_2013_QWise.pdf}}
% \begin{figure}[H]
% \centering
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{supervised_without_metric.png}
%   \caption{Original points, before metric learning}
%   \label{fig:sub1}
% \end{subfigure}%
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{supervised_with_metric.png}
%   \caption{points after metric learning (ITML)}
%   \label{fig:sub2}
% \end{subfigure}
% \caption{Metric learning with pairwise constraints}
% \label{fig:test}
% \end{figure}
% \paragraph{Weakly Supervised Metric Learning algorithms}
% Weakly Supervised algorithms use less information than the previous algorithms: they only take tuple of points and possibly a corresponding label. See below for more concrete examples of algorithms learning on tuples.
\item \textbf{Pair learners.} 
Another way is to have pair of points as inputs, and a corresponding label indicating whether the two points are similar or not. We can then learn a metric that brings points from a similar pair closer together, and points from a dissimilar pair further away from each other. In many applications, this can be a much easier type of supervision to collect, in particular when there are many labels. %, since annotators will only need to say whether images are similar or not, and will not have to assign a label in a huge list of those.
For instance, it is much easier for an annotator to decide whether two images correspond to the same person than to associate an individual image to the right identity among a large number of people (see Figure~\ref{fig:pairs}). 
% \begin{figure}[H]
% \centering
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{pairs_without_metric.png}
%   \caption{Original points, before metric learning}
%   \label{fig:sub1}
% \end{subfigure}%
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{pairs_with_metric.png}
%   \caption{points after metric learning (ITML)}
%   \label{fig:sub2}
% \end{subfigure}
% \caption{Metric learning with pairwise constraints}
% \label{fig:test}
% \end{figure}
\item \textbf{Quadruplet Learners.}
The last possibility to learn a metric in \texttt{metric-learn} is to take as input quadruplets of points. We will learn a metric that brings the two first points of each quadruplet closer than the two last points. This can be used for instance to learn a metric space in which closer points have closer values of a given attributes. This is very useful if the attribute is not binary like in pair supervision (e.g. same person/not same person), but is rather continuous and difficult to annotate precisely (e.g., the age of a person on an image, see Figure \ref{fig:quadruplets}).
%for instance, we want to learn a metric that puts closer together people from approximately the same age.
\end{itemize}

@@

\aurelien{mention optimization somewhere?}
\texttt{metric-learn} provides a unified interface to all metric learning algorithms (including the weakly supervised ones that have an input format that differs from the classic machine learning settings) and is designed to be fully compatible with the functionalities of \texttt{scikit-learn} \citep{scikit-learn}. All metric learners in \texttt{metric-learn} inherit from an abstract \texttt{BaseMetricLearner} object, which itself inherits from \texttt{BaseEstimator}. All classes inheriting from \texttt{BaseMetricLearner} should implement two methods, \texttt{get\_metric} and \texttt{score\_pairs}: the former returns a function that computes the distance between two points $u$ and $v$. The latter method computes the distances between a dataset of pairs of points.
\aurelien{score\_pairs and get\_metric should be mentioned here (all metric learners should implement these). In Mahalanobis, just mention they are implemented accordingly (do not delete: TODO: update accordingly mahalanobis}
% \todo[inline]{Should I talk about the classes etc ? Not useful since they are not visible by the user right ?}
% \todo[inline]{I think it can be nice, this is not documentation, but explaining how things are coded}

@@

\paragraph{Weakly Supervised metric learners} 
% Algorithms that learn on pairs or quadruplets take as input 3D array of shape \texttt{(n\_samples, t, n\_features)} (where...). In other words, sampling one element along the first dimension returns one tuple (e.g. a pair), that is a matrix of t lines corresponding to the t elements in the tuple.
% This format is what allows all these API choices, since slicing along the 1st dimension
% While a classical input array \texttt{X} in \texttt{scikit-learn} has shape \texttt{(n\_samples, n\_features)}, an array of tuples has shape \texttt{(n\_tuples, t, n\_features)}, where \texttt{t} is the number of elements in a tuple (e.g. 2 for pairs).
% Note that \texttt{scikit-learn} accepts arrays in a broad sense, in what they call "array-like" objects. \texttt{metric-learn} too accepts "tuple-like" objects.
Weakly supervised learning algorithms are similar to \texttt{scikit-learn} classifiers, but they \texttt{fit} on tuples that can be given as 3D arrays, instead of 2D arrays-like of points \texttt{X}. They have a \texttt{predict} method, that gives a prediction for each given tuple (e.g. for pairs it predicts whether the two points in a pair or similar or not). Tuples can be pairs or quadruplets, depending on the specific algorithm used. (see section~\ref{metriclearning}). Say we work with $n_t$ tuples of points with $p$ features each.\\
\textit{(i) Pairs learners} take as input an array-like \texttt{pairs} of shape $(n_t, 2, p)$, as well as an array-like \texttt{y\_pairs} of shape $(n_t,)$, which for each pair tells whether it contains similar or dissimilar samples. In order to \texttt{predict} on incoming pairs whether they are similar or not, they also need to fit a threshold, i.e. a reference distance with respect to which the distance between the incoming points will be compared. If this distance is higher than the threshold, the pair will be predicted as dissimilar, and if smaller as similar. This threshold is automatically fitted at fit time using a heuristic, but can be set manually using the method \texttt{set\_threshold}, or calibrated more specifically on a validation set, in order to optimize any classification score, using the method \texttt{calibrate\_threshold}.\\
\textit{(ii) Quadruplets learners} work on arrays-like of shape $(n_t, 4, p)$, and no labels array, where for each quadruplet the two first elements are the ones we want to be closer than the two last ones. They can also \texttt{predict} whether an incoming quadruplet is in the right order or not, i.e. whether the two first elements are indeed more similar that the two last, by comparing the two pairwise distances.
\aurelien{a bit more details would be welcome (explain both case of pairs and quadruplets). for pairs, need to mention threshold and calibration.}

>>>>

Metric learning algorithms can be classified according to the form of data supervision that they need to learn the metric. \texttt{metric-learn} currently implements algorithms that fall into the following categories:
\begin{itemize}
\item \textbf{Supervised learners} learn from a standard labeled dataset and essentially aim to bring closer together points from the same class while spreading further away points from a different class. For instance, the class may be the identity of a person in an image (see Figure~\ref{fig:full}). 
% \todo[inline]{ces figures ne sont pas très vendeuses. je propose de faire une figure conceptuelle en prenant l'exemple des faces expliquant la différence entre le supervisé, les paires et les quadruplets. on peut s'inspirer de \url{http://cedric.cnam.fr/~thomen/papers/Law_ICCV_2013_QWise.pdf}}
% \begin{figure}[H]
% \centering
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{supervised_without_metric.png}
%   \caption{Original points, before metric learning}
%   \label{fig:sub1}
% \end{subfigure}%
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{supervised_with_metric.png}
%   \caption{points after metric learning (ITML)}
%   \label{fig:sub2}
% \end{subfigure}
% \caption{Metric learning with pairwise constraints}
% \label{fig:test}
% \end{figure}
% \paragraph{Weakly Supervised Metric Learning algorithms}
% Weakly Supervised algorithms use less information than the previous algorithms: they only take tuple of points and possibly a corresponding label. See below for more concrete examples of algorithms learning on tuples.
\item \textbf{Pair learners} take as input a set of pairs of points with a pair label indicating whether the two points are similar or not. They aim to learn a metric that brings points from a similar pair closer together, and points from a dissimilar pair further away from each other. Such weak suIn many applications, this weaker supervision can be can be an easier type of supervision to collect, in particular when there are many labels. %, since annotators will only need to say whether images are similar or not, and will not have to assign a label in a huge list of those.
For instance, it is much easier for an annotator to decide whether two face images correspond to the same person than to match a face to its identity among a large number of people (see Figure~\ref{fig:pairs}). 
% \begin{figure}[H]
% \centering
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{pairs_without_metric.png}
%   \caption{Original points, before metric learning}
%   \label{fig:sub1}
% \end{subfigure}%
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.8\linewidth]{pairs_with_metric.png}
%   \caption{points after metric learning (ITML)}
%   \label{fig:sub2}
% \end{subfigure}
% \caption{Metric learning with pairwise constraints}
% \label{fig:test}
% \end{figure}
\item \textbf{Quadruplet Learners.}
The last possibility to learn a metric in \texttt{metric-learn} is to take as input quadruplets of points. We will learn a metric that brings the two first points of each quadruplet closer than the two last points. This can be used for instance to learn a metric space in which closer points have closer values of a given attributes. This is very useful if the attribute is not binary like in pair supervision (e.g. same person/not same person), but is rather continuous and difficult to annotate precisely (e.g., the age of a person on an image, see Figure \ref{fig:quadruplets}).
%for instance, we want to learn a metric that puts closer together people from approximately the same age.
\end{itemize}

@@

\aurelien{mention optimization somewhere?}
\texttt{metric-learn} provides a unified interface to all metric learning algorithms (including the weakly supervised ones that have an input format that differs from the classic machine learning settings) and is designed to be fully compatible with the functionalities of \texttt{scikit-learn} \citep{scikit-learn}. All metric learners in \texttt{metric-learn} inherit from an abstract \texttt{BaseMetricLearner} object, which itself inherits from scikit-learn's \texttt{BaseEstimator}. All classes inheriting from \texttt{BaseMetricLearner} should implement two methods, \texttt{get\_metric} and \texttt{score\_pairs}: the former returns a function that computes the distance between two points $u$ and $v$. The latter method computes the distances between a dataset of pairs of points.
\aurelien{score\_pairs and get\_metric should be mentioned here (all metric learners should implement these). In Mahalanobis, just mention they are implemented accordingly (do not delete: TODO: update accordingly mahalanobis}
% \todo[inline]{Should I talk about the classes etc ? Not useful since they are not visible by the user right ?}
% \todo[inline]{I think it can be nice, this is not documentation, but explaining how things are coded}

@@

\william{cf. Nathalie's comment: re-introduce weakly supervised metric laerners}
\paragraph{Weakly Supervised metric learners} 
% Algorithms that learn on pairs or quadruplets take as input 3D array of shape \texttt{(n\_samples, t, n\_features)} (where...). In other words, sampling one element along the first dimension returns one tuple (e.g. a pair), that is a matrix of t lines corresponding to the t elements in the tuple.
% This format is what allows all these API choices, since slicing along the 1st dimension
% While a classical input array \texttt{X} in \texttt{scikit-learn} has shape \texttt{(n\_samples, n\_features)}, an array of tuples has shape \texttt{(n\_tuples, t, n\_features)}, where \texttt{t} is the number of elements in a tuple (e.g. 2 for pairs).
% Note that \texttt{scikit-learn} accepts arrays in a broad sense, in what they call "array-like" objects. \texttt{metric-learn} too accepts "tuple-like" objects.
Weakly supervised learning algorithms are similar to \texttt{scikit-learn} classifiers, but they \texttt{fit} on tuples that can be given as 3D arrays, instead of 2D arrays-like of points \texttt{X}. They have a \texttt{predict} method, that gives a prediction for each given tuple (e.g. for pairs it predicts whether the two points in a pair or similar or not). Tuples can be pairs or quadruplets, depending on the specific algorithm used. (see section~\ref{metriclearning}). Say we work with $n_t$ tuples of points with $p$ features each.\\
\textit{(i) Pairs learners} take as input an array-like \texttt{pairs} of shape $(n_t, 2, p)$, as well as an array-like \texttt{y\_pairs} of shape $(n_t,)$, which for each pair tells whether it contains similar or dissimilar samples. In order to \texttt{predict} on incoming pairs whether they are similar or not, they also need to fit a threshold, i.e. a reference distance with respect to which the distance between the incoming points will be compared. If this distance is higher than the threshold, the pair will be predicted as dissimilar, and if smaller as similar. This threshold is automatically fitted at fit time using a heuristic, but can be set manually using the method \texttt{set\_threshold}, or calibrated more specifically on a validation set, in order to optimize any classification score, using the method \texttt{calibrate\_threshold}.\\
\textit{(ii) Quadruplets learners} work on arrays-like of shape $(n_t, 4, p)$, and no labels array, where for each quadruplet the two first elements are the ones we want to be closer than the two last ones. They can also \texttt{predict} whether an incoming quadruplet is in the right order or not, i.e. whether the two first elements are indeed more similar that the two last, by comparing the two pairwise distances.
\aurelien{a bit more details would be welcome (explain both case of pairs and quadruplets). for pairs, need to mention threshold and calibration.}
